---
title: "Does beaver damming influence arthropods' individual size distribution?"
author: 
- name: Leonardo Capitani 
  email: leonardo.capitani@eawag.ch
date: "10/01/2025"
date-modified: last-modified
execute:
  echo: true
  warning: false
  message: false
format: 
      html:
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
      pdf:
             keep-tex: true
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: eco_lett.csl
---

```{r loadlib}
library(tidyverse)
library(DT)
library(here)
library(dplyr)
library(poweRlaw)
library(isdbayes)
library(brms)
library(tidybayes)
```

```{r ggplot_theme}
newtheme <- theme_bw(base_size = 14) 
theme_set(newtheme)
```

## Load data

We are going to load the kick-net data set.

```{r load_data_k}
k = read.csv(here("data", "processed", "kicknet_clean.csv"))
```

### Standardize human influence

```{r stardize_k2}

k2 <- k  |>  
  mutate(across(c(human_influence), 
                ~ .x %>% scale() %>% as.vector(), 
                .names = "{.col}_z"),
         ecomorphology = as.factor(ecomorphology)) # ecomorphology as factor 
```

Then we add the column count:

```{r k3}
k3 <- k2 |> 
  mutate(counts = 1)
```

Now I am going to add the `xmin` and `xmax` columns by `site` and `location`:

```{r xmin_xmax}
k4 <-  k3 |> group_by(location, site)  |> 
  mutate(
    xmin = min(dw, na.rm = TRUE),
    xmax = max(dw, na.rm = TRUE)
  )  |> 
  ungroup()
```

See the values in a dynamic table:

```{r t}
t <- k3  |> 
  group_by(location, site)  |> 
  summarise(
    xmin = round(min(dw, na.rm = TRUE),3),
    xmax = round(max(dw, na.rm = TRUE),3),
    .groups = "drop")

datatable(t)
```

## Addressing Undersampling

In field collections of body sizes, small organisms are commonly under-sampled. This contradicts that assumption of an exponential decline in frequency as body size increases and can lead to dramatically poor estimates of lambda [@clauset2009]. For this reason, [Jeff Wesner](jeff.wesner@usd.edu) and [Justin Pomeranz](jfpomeranz@gmail.com) strongly recommend checking for and removing under-sampled small sizes.

### Find and remove biased sizes

How do we know if body sizes are under-sampled? One option is to use the `estimate_xmin()` function from the `poweRlaw` package [@gillespie2015] (to install the package read [here](https://github.com/csgillespie/poweRlaw)). It first fits the data to a power law (unbounded, but that’s OK here). Then it uses Kolmogorov-Smirnoff sampling to find the minimum body mass above which the data follow a power law.

::: callout-warning
Jeff Wesner:

This trimming procedure is sample-specific, not global. So if you have multiple samples, then you’ll need to run the `estimate_xmin` procedure on each sample, trim with a unique `xmin_est`, and recalculate xmin for each sample.
:::

```{r estimate_xmin}
removal_summary <- k4 %>%
  group_by(site, location) %>%
  summarise(
    # Fit continuous power law and estimate xmin
    xmin_est = round({
      temp_mod <- conpl$new(dw)
      estimate_xmin(temp_mod)$xmin
    }, 2),
    # Total number of observations in this group
    total_obs = n(),
    # Number of observations below xmin (to be removed)
    removed_obs = sum(dw < xmin_est),
    # Proportion of observations below xmin
    prop_removed = round(removed_obs / total_obs * 100, 1),
    # Remaining observations after culling
    remaining_obs = total_obs - removed_obs,
    .groups = "drop"
  )

# dynamic data table to inspect
datatable(
  removal_summary,
  options = list(
    pageLength = 10,   # rows per page
    autoWidth = TRUE,  # auto adjust column widths
    searching = TRUE,  # allow search/filter
    ordering = TRUE    # allow sorting
  ),
  rownames = FALSE,
  caption = "Observations Removed per Site-Location"
)

```

> Jeff Wesner: note that the trimming can be quite large despite the relatively small `xmin_est` value. It can sometimes remove much more (such as 25 or 50%). That is usually not a problem as long as the remaining data has enough individuals to fit reliably, usually 300 or so.

::: callout-note
Question: What if we group only by `location` ? Since `site` will be consider as random effect ... or not consider at all.. maybe we can assume that all individuals that were collected in the, for example, `control` location, follows the same Pareto distribution.
:::

If we go on with this approach, then we are going to loose for the kick net method:

```{r trigger}
# --- Join back to remove observations ---
k5 <- k4 %>%
  left_join(removal_summary %>% select(site, location, xmin_est),
            by = c("site", "location")) %>%
  filter(dw >= xmin_est) %>%       # remove observations below xmin
  select(-xmin_est)  |>               # drop helper column if not needed
group_by(location, site)  |> # recompute the xmin and xmax
  mutate(
    xmin = min(dw, na.rm = TRUE),
    xmax = max(dw, na.rm = TRUE)
  )  |> 
  ungroup()
```

::: callout-important
We trigger `r nrow(k4) - nrow(k5)` observations (`r round((nrow(k4) - nrow(k5)) / nrow(k4) * 100, 1)`%).

Remaining observations are:\
`r paste(k5 %>% count(location) %>% mutate(txt = paste0(location, " = ", n)) %>% pull(txt), collapse = "; ")`.
:::

## Preliminary data visualization

Q: Does ecomorphology vary by site and by location ?

```{r variation_ecomorphology}

ggplot(k2, aes( x = site, y = ecomorphology))+
  geom_point()+
  theme_bw(base_size = 14)+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
ggplot(k2, aes( x = location, y = ecomorphology))+
  geom_point()+
  theme_bw(base_size = 14)+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

Ecomorphology vary by site and by location.

Does human influence vary by site and location ?

```{r human_var_sentences}

sentences <- k2 %>%
  mutate(human_influence_z = round(human_influence_z, 2)) %>%
  group_by(site) %>%
  summarise(
    all_equal = n_distinct(human_influence_z) == 1,
    details = if (n_distinct(human_influence_z) == 1) {
      paste0("At site ", first(site), 
             ", all locations have the same human_influence_z value of ", 
             first(human_influence_z), ".")
    } else {
      paste0("At site ", site, " and location ", location, 
             ", the human_influence_z value is ", human_influence_z, ".",
             collapse = " ")
    },
    .groups = "drop"
  ) %>%
  pull(details)

sentences


```

```{r human_influence_variation}

ggplot(k2, aes( x = site, y = human_influence_z, color = location)) + 
  geom_point()+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Human influence does not vary by site.

::: callout-note
Which distribution does the data look like?
:::

```{r hist, fig.width=10, fig.height=12}
# Get unique sites
sites_list <- unique(k4$site)

# First 8 sites
k4_first8 <- k4 %>% filter(site %in% sites_list[1:8])

ggplot(k4_first8, aes(x = dw)) +
  geom_histogram(
    binwidth = 0.1,
    color = "black", 
    fill = "darkgrey"
  ) +
  facet_grid(site ~ location, scales = "free_y") +
  scale_x_continuous(limits = c(0,20)) +
  labs(
    title = "Kick net method (Sites 1–8)",
    x = "Dry weight (mg)",
    y = "Count", 
    subtitle = "Small invertebrate body sizes undersampled"
  ) 

# Next 8 sites
k4_next8 <- k4 %>% filter(site %in% sites_list[9:16])

ggplot(k4_next8, aes(x = dw)) +
  geom_histogram(
    binwidth = 0.1,
    color = "black", 
    fill = "darkgrey", 
    alpha = 0.7
  ) +
  facet_grid(site ~ location, scales = "free_y") +
  scale_x_continuous(limits = c(0,20)) +
  labs(
    title = "Kick net method (Sites 9–16)",
    x = "Dry weight (mg)",
    y = "Count", 
    subtitle = "Small invertebrate body sizes undersampled"
  )

```

Then we plot the triggered data frame:

```{r hist2, fig.width=10, fig.height=12}
# Get unique sites
sites_list <- unique(k5$site)

# First 8 sites
k5_first8 <- k5 %>% filter(site %in% sites_list[1:8])

ggplot(k5_first8, aes(x = dw)) +
  geom_histogram(
    binwidth = 0.25,
    #color = "white", 
    fill = "steelblue",
  ) +
  facet_grid(site ~ location, scales = "free_y") +
  scale_x_continuous(limits = c(0,20)) +
  labs(
    title = "Kick net method (Sites 1–8)",
    x = "Dry weight (mg)",
    y = "Count", 
    subtitle = "Small invertebrate body sizes removed"
  ) 

# Next 8 sites
k5_next8 <- k5 %>% filter(site %in% sites_list[9:16])

ggplot(k5_next8, aes(x = dw)) +
  geom_histogram(
    binwidth = 0.25,
    #color = "white", 
    fill = "steelblue", 
    alpha = 0.7
  ) +
  facet_grid(site ~ location, scales = "free_y") +
  scale_x_continuous(limits = c(0,20)) +
  labs(
    title = "Kick net method (Sites 9–16)",
    x = "Dry weight (mg)",
    y = "Count", 
    subtitle = "Small invertebrate body sizes removed"
  )

```

## Model definition

First, the model formula:

```{r model_formula}

# formula 
form = bf(dw | vreal(counts, xmin, xmax) ~ 0  + location, # 0 + allows to see the four locations with its own lambda . if you do x ~ 1  + location, then control location will consider your reference level (intercept).
           family = paretocounts())

```

### Default priors

Second, which default priors are we using:

```{r check_priorss}

# see -the default / improper priors that the mdoel will use 
default_prior(object = form, data = k5 )
```

Third, let's set our own priors:

```{r setpriors}
#### set priors ####

pr  <- c( # prior for the average (intercept) of predictors 
  set_prior("normal(-1.20, 0.5)", coef = "locationControl"), # this is lambda for control location! )
  set_prior("normal(-1.20, 0.5)", coef = "locationOutflow"),
  set_prior("normal(-1.20, 0.5)", coef = "locationPool"),
  set_prior("normal(-1.20, 0.5)", coef = "locationInflow")
  ) 

```

Let's validade if the priors are retained by the model without errors:

```{r validapriors}
### validate priors: check if we have correctly state the priors

validate_prior(prior = pr, formula = form, data = k5 )
```

### Fit the model 

Let's see if the model is going to work:

```{r fit_the_model}

m1 = brm(formula = form , 
           data = k5,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 4, 
           iter = 1000,
           seed = 7, # allow exact replication of results 
          file = here("results","models","m1_removed"), # save the model. Please set your path folder. Note that if you want to refit the model, then you need to hide/deactivate this line of code 
        sample_prior = "yes" )# options: "yes", "no" or "only". If you want to sample from priors, use ONLY. 
```

Get the summary of the model:

```{r summarzm1}
summary(m1)
```

Plot posterior distributions using `add_epred_draws` or `add_predicted_draws`

```{r pp_epred}
pp = m1 |> 
   tidybayes::add_epred_draws(seed = 9, # assure reproducibility  
                      ndraws = 500,
                      newdata = k5)

pp_sigma = m1 |> 
   tidybayes::add_predicted_draws(seed = 9, # assure reproducibility  
                      ndraws = 500,
                      newdata = k5)
```

```{r ggplot_epred}

ggplot(data = pp, aes(y = location, x = .epred, fill = location)) + 
  stat_halfeye(scale = 0.5) +
  labs(x = expression(lambda),
     y = "location", 
     fill = "location")+
  theme_bw(base_size = 14)
```

As far as I understand, in a multilevel model, there are three sources of uncertainty:

1.  the residual (observation-level) variance ($\sigma$)
2.  the uncertainty in the fixed coefficients
3.  the uncertainty in the variance parameters for the grouping factors

To include all sources of uncertainty call ``` brms::``predict(.) ``` or ``` tidybayes::``add_predicted_draws(.) ``` or `tidybayes::predicted_draws(.)` or `brms::posterior_predict(.)`.

To exclude 1. call `fitted(.)` or `brms::posterior_linepred(.)` or `linpred_draws(.)` or `tidybayes::add_fitted_draws(.)`.

To exclude 3. (possible partially if you want), use the `re_formula` argument.

The results from `posterior_predict(.)` and `posterior_linpred(.)` have the same mean, but the full posterior predictions that incorporate the estimated $\sigma$ have a much wider range of plausible values.

Special mention to:

The results from [`posterior_epred()`](https://mc-stan.org/rstantools/reference/posterior_epred.html) are a little strange to understand, and in the case of normal/Gaussian regression (and many other types of regression models!), they’re identical to the linear predictor ([`posterior_linpred()`](https://mc-stan.org/rstantools/reference/posterior_linpred.html)). These are the posterior draws of the expected value or mean of the the posterior distribution, or in the model. Behind the scenes, this is calculated by taking the average of each row’s posterior distribution and then taking the average of *that*.

### Prior vs posterior

```{r tidy_prior_vs_posterior}
# Let's also visualize them. 
library(bayesplot)
mcmc_intervals(m1)


tidy_inter <- bayesplot::mcmc_intervals_data(m1, prob = 0.66, # 66% intervals
                                  prob_outer = 0.95,
                                  #transformations =  "exp",
                                  pars = vars(contains(c("prior_", "sigma","b_", "bsp","nu")) & -contains(c("sd_", "lp__", "lprior"))))


tidy_inter1 <- tidy_inter %>%
  mutate(type = ifelse(grepl("^prior", sub("_.*", "", parameter)), "Prior", "Posterior"),
         parameter = gsub("^prior_", "", parameter) ) 



```

## Replicate the vignettes for `isdbayes`

You can see the original vignettes [here](https://github.com/jswesner/isdbayes).

First, simulate some power law data using `rparetocounts()`. The code below simulates 300 body sizes from a power law with exponent lambda = -1.2, xmin = 1, and xmax = 1000.

```{r simdata}

# simulate data

dat = tibble(dw = rparetocounts(n = 300,  lambda = -1.2,  xmin = 1, xmax = 1000)) |> 
  mutate(xmin = min(dw),
         xmax = max(dw),
         counts = 1)
```

## Model fit

Next estimate the power law exponent using `brms`. The model below (`fit1`) is an intercept only model, where x are the body sizes and counts, xmin, and xmax are included in `vreal()`. The use of `vreal` has nothing to do with the model per se. It is simply required wording from `brms` when including custom families. Similarly, `stanvars` is required wording that contains the custom likelihood parameters. As long as `isdbayes` is loaded, then `stanvars = stanvars` will work. It will stay the same regardless of changes to the model structure (like new predictors or varying intercepts).

```{r set_formula}
# formula 
bform = bf(dw | vreal(counts, xmin, xmax) ~ 0 + Intercept, # 0 + Intercept is the same as ~ 1, but the first is an explicit way to code the intercept. While the second is implicit. In this case the intercept is our lambda 
           family = paretocounts())

# see the default / improper priors that the mdoel will use 
default_prior(object = bform, data = dat )
#### set priors ####

pr  <- c( # prior for the average (intercept) of predictors 
  set_prior("normal(-1.20, 0.5)", coef = "Intercept") # this is lambda ! )
  ) 
### validate priors: check if we have correctly state the priors

validate_prior(prior = pr, formula = bform, data = dat )

# yes ! Now our model is expecting to find lamba values in between -1.2 +- 0.5 
```

```{r fit_model}

fit0 = brm(formula = bform, 
          data = dat,
          stanvars = stanvars,    # required for truncated Pareto
          family = paretocounts(),# required for truncated Pareto
          prior = pr, # here we set the priors 
          chains = 3, 
          warmup = 500, 
          iter = 1000, 
          seed = 7, # allow exact replication of results 
          file = here("results","models","fit0"), # save the model. Please set your path folder. Note that if you want to refit the model, then you need to hide/deactivate this line of code 
        sample_prior = "no"  # default option. Indicate if draws from priors should be drawn additionally to the posterior draws.
          )
```

```{r summary}

summary(fit0)
```

Plot posterior distribution of lambda:

```{r posterior_draws}
p = fit0 |> 
   add_epred_draws(seed = 9, # assure reproducibility  
                      ndraws = 1000,
                      newdata = dat)
```

```{r plot_lambda}
p |> 
  ggplot(aes(x = .epred, y = 0)) + 
  stat_halfeye() + 
  labs(x = expression(lambda))+
  theme_bw(base_size = 14)+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

## Fit multiple size distributions with a fixed factor

Now we are going to simulate data sampled in four different `locations.`

```{r simdata_multilevels}

control = rparetocounts(lambda = -1.8) # `lambda` is required wording from brms. in this case it means the lambda exponent of the ISD
inflow = rparetocounts(lambda = -1.5)
pool = rparetocounts(lambda = -1.2)
outflow = rparetocounts(lambda = -1.1)

isd_data = tibble(control = control,
                  inflow = inflow,
                  pool = pool,
                  outflow = outflow) |> 
  pivot_longer(cols = everything(), names_to = "group", values_to = "dw") |> 
  group_by(group) |> 
  mutate(xmin = min(dw),
         xmax = max(dw)) |> 
  group_by(group, dw) |> 
  add_count(name = "counts")
```

Set formula and priors:

```{r setmultilevel_formula}

# formula 
bform2 = bf(dw | vreal(counts, xmin, xmax) ~ 0  + group, # 0 + allows to see the four locations with its own lambda . if you do x ~ 1  + group, then control location will consider your reference level. Try by yourself . 
           family = paretocounts())
```

### Set the priors

```{r set_priors}
# see the default / improper priors that the mdoel will use 
default_prior(object = bform2, data = isd_data )
#### set priors ####

pr  <- c( # prior for the average (intercept) of predictors 
  set_prior("normal(-1.20, 0.5)", coef = "groupcontrol"), # this is lambda for control location! )
  set_prior("normal(-1.20, 0.5)", coef = "groupinflow"),
  set_prior("normal(-1.20, 0.5)", coef = "grouppool"),
  set_prior("normal(-1.20, 0.5)", coef = "groupoutflow")
  ) 
### validate priors: check if we have correctly state the priors

validate_prior(prior = pr, formula = bform2, data = isd_data )

# yes ! Now our model is expecting to find lamba values in between -1.2 +- 0.5 
```

### Fit the model 

```{r fit2}

fit2 = brm(formula = bform2 , 
           data = isd_data,
           stanvars = stanvars,
           family = paretocounts(),
           chains = 4, 
           iter = 1000,
           seed = 7, # allow exact replication of results 
          file = here("results","models","fit2"), # save the model. Please set your path folder. Note that if you want to refit the model, then you need to hide/deactivate this line of code 
        sample_prior = "no" )
```

Plot posterior distributions, a.k.a:

```{r posterior_draws2}

p2 = fit2 |> 
   add_epred_draws(seed = 9, # assure reproducibility  
                      ndraws = 1000,
                      newdata = isd_data)
```

Plot the lambdas

```{r }

# data frame of known lambda values
lambda_vals <- data.frame(
  group = c("control", "inflow", "pool", "outflow"),
  lambda = c(-1.8, -1.5, -1.2, -1.1)
)


ggplot(data = p2, aes(y = group, x = .epred, fill = group)) + 
  stat_halfeye(scale = 0.5) + 
  # add stars at the right group + lambda
  geom_point(
    data = lambda_vals,
    aes(x = lambda, y = group),
    shape = 8,   # star
    size = 4,
    color = "red"
  ) +
  labs(x = expression(lambda),
     y = "location", 
     fill = "location",
     subtitle = "Red stars represent the a priori simulated value of lambdas")+
  theme_bw(base_size = 14)+ 
  scale_x_continuous(breaks = seq(-2, -1, by = 0.25), limits = c(-2,-1))
```

## Visualize ISD

> Jeff Wesner: This code extracts the cumulative probabilities using `pparetocounts()` and the plots them over raw data. Note that the raw data probabilities are simply estimates for visualization purposes. These plots are typical in studies of the ISD and are visually similar to plots of log-abundance vs log-size, making them more familiar to readers (maybe).

```{r visualize_isd}


# 1) sort raw data
d = fit0$data |>
  arrange(-dw) |> 
  mutate(order = row_number(),
         y_raw_prob = order/max(order)) # convert to 0-1 scale

# 2) data grid to sample over
data_grid = d |>
  distinct(xmin, xmax) |> 
  expand_grid(dw = 2^seq(log2(min(d$dw)), log2(max(d$dw)), length.out = 30)) |> # sequence is log 2 to ensure equal logarithmic spacing
  mutate(counts = 1) # This is a default. Even if counts are >1 inthe raw data, make them = 1 here.

# 3) extract posteriors
isd_posts = data_grid |> 
  tidybayes::add_epred_draws(fit0)

# 4) get cumulative probabilities from posteriors
isd_lines = isd_posts |> 
  rowwise() |> 
  mutate(y_prob = pparetocounts(x = dw, xmin = xmin ,xmax = xmax, lambda = .epred))

# 5) plot raw vs posterior samples
isd_lines |> 
  filter(.draw <= 100) |> # limits to the first 100 draws. Change as needed or use a summary to plot instead of individual lines
  ggplot(aes(x = dw, y = y_prob)) + 
  geom_line(aes(group = .draw), alpha = 0.3) +
  geom_point(data = d, aes(y = y_raw_prob),shape = 21, fill = "white", color = "black") +
  labs(y = "P(X >= x)")+
  theme_bw(base_size = 14)
```

## References
